---
title: "SE 412 Project 1 - Market Basket Analysis"
author: "Khusbu Bhushal"
date: "2025-11-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# INTRODUCTION
This project uses market basket analysis to explore frequent itemsets and association rules.  

# LOAD PACKAGES
```{r packages, message=FALSE, warning=FALSE}
library(arules)
library(arulesViz)
library(tidyverse)
library(readxl)
library(plyr)
library(knitr)
```

# LOAD DATA
```{r load-data}
Itemslist <- read.csv2("Assignment-1_Data.csv", check.names = FALSE)
Itemslist <- Itemslist[complete.cases(Itemslist), ]
head(Itemslist)
```

# PREPROCESSING
```{r preprocessing}
# Group items into full transactions
transactionData <- ddply(Itemslist, c("BillNo","Country"),
                         function(df1) paste(df1$Itemname, collapse=","))

colnames(transactionData) <- c("BillNo", "Country", "items")

# Create France & Other datasets
franceData <- subset(transactionData, Country == "France")
otherData  <- subset(transactionData, Country != "France")

# Remove Country column
franceData$Country <- NULL
otherData$Country  <- NULL

colnames(franceData) <- c("BillNo", "items")
colnames(otherData)  <- c("BillNo", "items")

# Prepare item-only data frames for Apriori
transactionData_apriori <- data.frame(items = transactionData$items)
france_apriori          <- data.frame(items = franceData$items)
other_apriori           <- data.frame(items = otherData$items)

# Save CSVs
write.csv(transactionData_apriori, "transactions_basket.csv", quote=FALSE, row.names=FALSE)
write.csv(france_apriori,          "france_basket.csv",       quote=FALSE, row.names=FALSE)
write.csv(other_apriori,           "other_basket.csv",        quote=FALSE, row.names=FALSE)

# Load into transaction objects
trans_all   <- read.transactions("transactions_basket.csv", format="basket", sep=",")
trans_fra   <- read.transactions("france_basket.csv",       format="basket", sep=",")
trans_other <- read.transactions("other_basket.csv",        format="basket", sep=",")

summary(trans_all)
summary(trans_fra)
summary(trans_other)
```

##Everthing above in one dataframe is a tiny 3 row output that shows the printed example of item labels insidet the transaction object
##its purpose is to show a preview of item names so so you know what type of items are inside the transactions
##The bigger R Console data frame like output is a transaction length distribution that says:
## 1521 transactions contain 1 item , 843 transactions contain 2 items , 735 transactions contain 3 items , 419 items is the longest basket in the dataset (very large transaction)

---

# RQ1: What frequent itemsets and rules appear in the FULL dataset?

## RUN APRIORI FOR FULL DATASET
```{r apriori-full}
rules_all <- apriori(trans_all,
                     parameter = list(supp = 0.01, conf = 0.5))

summary(rules_all)
inspect(head(rules_all, 10))
```
### The Apriori algorithm discovered several very strong association rules in the full dataset. For example, {SET 3 RETROSPOT TEA} ⇒ {SUGAR} has 100% confidence and lift = 86.49, indicating that customers who purchase this tea set are extremely likely to also purchase sugar. The reversed rule {SUGAR} ⇒ {SET 3 RETROSPOT TEA} also holds with the same confidence and lift, meaning these items almost always appear together. These high-lift rules show strong product affinities that could inform product placement and bundling strategies.
### IT also shows set of 159 rules
###Most rules contained 2–3 items, showing that customers tend to purchase small item combinations
###Support

###→ Rules appear in 1%–2% of all transactions.

###Confidence
###→ Many rules are strong because confidence is 0.50 to 1.00
###(50%–100% probability of RHS given LHS)

###Lift
###→ Lift values range from 6 to 86, meaning:
###lift > 1 = items occur together more than random chance
###lift > 10 = very strong association
###lift ~ 80 = extremely strong association


## ITEM FREQUENCY PLOT
```{r item-frequency-plot, fig.width=8, fig.height=6}
itemFrequencyPlot(trans_all,
                  topN = 20,
                  type = "absolute",
                  main = "Top 20 Most Frequent Items")
```

## SCATTER PLOT FOR ALL RULES
```{r scatter-all, fig.width=8, fig.height=6}
plot(rules_all,
     measure = c("support","confidence"),
     shading = "lift",
     main = "Support vs Confidence (Colored by Lift)")
```

## GRAPH OF TOP 20 STRONGEST RULES
```{r graph-top20, fig.width=8, fig.height=6}
top20 <- head(sort(rules_all, by="lift"), 20)
plot(top20, method="graph", control=list(type="items"),
     main="Top 20 Strongest Association Rules")
```

---

# RQ2: How do rules differ for France vs Other Regions?

## APRIORI FOR FRANCE
```{r apriori-france}
rules_fra <- apriori(trans_fra,
                     parameter = list(supp = 0.01, conf = 0.5))
summary(rules_fra)
inspect(head(rules_fra, 10))
```

## APRIORI FOR OTHER REGIONS
```{r apriori-other}
rules_other <- apriori(trans_other,
                       parameter = list(supp = 0.01, conf = 0.5))
summary(rules_other)
inspect(head(rules_other, 10))
```

## SCATTER PLOT - FRANCE
```{r plot-france, fig.width=8, fig.height=6}
plot(rules_fra,
     measure = c("support","confidence"),
     shading = "lift",
     main="France: Support vs Confidence")
```

## SCATTER PLOT - OTHER REGIONS
```{r plot-other, fig.width=8, fig.height=6}
plot(rules_other,
     measure = c("support","confidence"),
     shading = "lift",
     main="Other Regions: Support vs Confidence")
```

## GRAPH OF STRONGEST RULES - FRANCE
```{r graph-france, fig.width=8, fig.height=6}
top_fra <- head(sort(rules_fra, by="lift"), 10)
plot(top_fra, method="graph", control=list(type="items"),
     main="France: Top 10 Strongest Rules")
```

## GRAPH OF STRONGEST RULES - OTHER REGIONS
```{r graph-other, fig.width=8, fig.height=6}
top_other <- head(sort(rules_other, by="lift"), 10)
plot(top_other, method="graph", control=list(type="items"),
     main="Other Countries: Top 10 Strongest Rules")
```
